---
title: "PhRMA & EFPIA R&D Analysis"
author: "Bena Smith"
output:
  html_document:
    
    df_print: paged
  pdf_document:
      extra_dependencies: 
        dcolumn: null
editor_options: 
  markdown: 
    wrap: 72
---

```         
includes:
  in_header: preamble.tex
  
```

```{r}
library(ggplot2)
library(plm)
library(lmtest)
library(sandwich)
library(janitor)
library(dplyr)
library(readxl)
library(writexl)
library(lubridate)
library(readr)
library(nlme)
library(stringr)
library(broom)
library(tidyr)
library(stargazer)
library(PerformanceAnalytics)
library(tseries) # to test stationarity


```

# Data Cleaning

# EFPIA/ Europe

## EFPIA R&D spending by country by year in euros

From yearly pdf reports 2004: **LINK NEEDED** 2005:
<https://www.efpia.eu/media/15485/the-pharmaceutical-industry-in-figures-edition-2007.pdf>
2006:
<https://www.efpia.eu/media/15486/the-pharmaceutical-industry-in-figures-edition-2008.pdf>
2007:
<https://www.pharmine.eu/wp-content/uploads/2014/04/The_Pharma_Industry_in_figures_Edition_2009-20080612-009-EN-v1.pdf>
2008:
<https://www.sfee.gr/wp-content/uploads/2015/05/Figures_2010_Final-20100611-001-EN-v1_0.pdf>
2009:
<https://www.efpia.eu/media/15489/the-pharmaceutical-industry-in-figures-edition-2011.pdf>
2010:
<https://www.sfee.gr/wp-content/uploads/2015/05/EFPIA_Figures_2012_Final.pdf>
Data for 2011 could not be found 2012:
<https://www.phar-in.eu/wp-content/uploads/2014/05/Figures_2014_Final.pdf>
2013:
<https://www.efpia.eu/media/25822/2015-the-pharmaceutical-industry-in-figures.pdf>
2014:
<https://www.efpia.eu/media/25055/the-pharmaceutical-industry-in-figures-june-2016.pdf>
2015:
<https://www.efpia.eu/media/219735/efpia-pharmafigures2017_statisticbroch_v04-final.pdf>
2016:
<https://www.efpia.eu/media/361960/efpia-pharmafigures2018_v07-hq.pdf>
2017:
<https://www.efpia.eu/media/412931/the-pharmaceutical-industry-in-figures-2019.pdf>
2018:
<https://www.efpia.eu/media/554521/efpia_pharmafigures_2020_web.pdf>
2019:
<https://www.efpia.eu/media/602709/the-pharmaceutical-industry-in-figures-2021.pdf>
2020:
<https://www.efpia.eu/media/637143/the-pharmaceutical-industry-in-figures-2022.pdf>
2021:
<https://www.efpia.eu/media/rm4kzdlx/the-pharmaceutical-industry-in-figures-2023.pdf>

```{r}
efpia_rd <- read.csv("Cleaned Data/Area Level R&D Spending/EFPIA R&D Spending by Country in EU.csv")
efpia_rd <- clean_names(efpia_rd)
efpia_rd
```

formatting

```{r}
efpia_rd$r_d_spending_millions_euros <- gsub(",", "", efpia_rd$r_d_spending_millions_euros)
efpia_rd$r_d_spending_millions_euros <- as.numeric(efpia_rd$r_d_spending_millions_euros)
efpia_rd
```

sum total EU by year

```{r}
efpia_rd_totals <- efpia_rd %>%
  group_by(year) %>%
  summarise(total_mil_euros = sum(r_d_spending_millions_euros, na.rm = TRUE))
efpia_rd_totals

```

Dollars to euros
<https://www.macrotrends.net/2548/euro-dollar-exchange-rate-historical-chart>

```{r}
usdtoeuroavgconversion <- read_xlsx("Cleaned Data/Area Level R&D Spending/USD to Euro Conversion Avg byt Year.xlsx")
usdtoeuroavgconversion
eurotodollars <- as.data.frame(cbind(usdtoeuroavgconversion$Year, 1/usdtoeuroavgconversion$`Average Closing Price`))
colnames(eurotodollars) <- c("year", "euros_to_dollar_conversion")
eurotodollars
```

multiply conversion rate to get r&d spending in dollars

```{r}
rd_conversion_merged_df <- merge(efpia_rd_totals, eurotodollars, by = "year")
rd_conversion_merged_df
efpia_rd_totals$total_rd_mil_dollars <- rd_conversion_merged_df$euros_to_dollar_conversion * rd_conversion_merged_df$total_mil_euros
efpia_rd_totals$annual_domestic_RD_spending_bil_dollars <- efpia_rd_totals$total_rd_mil_dollars /1000

efpia_rd_totals
```

## EU orphan drug market authorizations

EU Orphan Drugs were scraped from a pdf
<https://www.orpha.net/pdfs/orphacom/cahiers/docs/GB/list_of_orphan_drugs_in_europe.pdf>
it can be found in the the Original Data Folder

```{r}
eu_orphan_drugs <- read_xlsx("Cleaned Data/Area Level Orphan Authorizations/EU_Orphan_w_Company from https___www.orpha.net_orphacom_cahiers_docs_GB_list_of_orphan_drugs_in_europe.pdf.xlsx")
eu_orphan_drugs <- clean_names(eu_orphan_drugs)
eu_orphan_drugs
```

Check for duplicated rows

```{r}
eu_orphan_drugs[duplicated(eu_orphan_drugs),]

```

EU number of market authorizations for each drug

```{r}
eu_ma_freq <- eu_orphan_drugs %>%
  count(tradename) %>%
  arrange(desc(n))

colnames(eu_ma_freq) <- c("Generic Name", "Market Authorization Frequency")
eu_ma_freq$`Generic Name` <- tolower(eu_ma_freq$`Generic Name`)
eu_ma_freq

```

How many ODs with more than 1 authorization

```{r}
nrow(eu_ma_freq[eu_ma_freq$`Market Authorization Frequency` > 1, ])
```

```{r}
stargazer(eu_ma_freq[1:3,], title= "EU Market Authorization Frequencies for Orphan Drugs with Over 1 Market Authorization", rownames = FALSE, align=TRUE, summary=FALSE, type = "html", out="Final Figures Formatted/EU_OD_Multiple_MAs.htm")

```

Look at where there are multiple market authorizations for the same drug

```{r}
eu_orphan_drugs[eu_orphan_drugs$tradename == "CARBAGLU",]
eu_orphan_drugs[eu_orphan_drugs$tradename == "NEXAVAR",]
eu_orphan_drugs[eu_orphan_drugs$tradename == "SOLIRIS",]

```

Select only new drugs in the EU

```{r}
eu_orphan_drugs_new <- eu_orphan_drugs %>%
  group_by(tradename) %>%  
  slice_min(order_by = market_authorization_year, n = 1, with_ties = FALSE) %>%  # Select the entry with the earliest market authorization year
  ungroup() 

eu_orphan_drugs_new
```

format new EU orphan drugs by year

```{r}
eu_new_od_ma_freqs <- as.data.frame(table(eu_orphan_drugs_new$market_authorization_year))
colnames(eu_new_od_ma_freqs) <- c("Year", "Freq New Orphan Drug MAs")
eu_new_od_ma_freqs

eu_new_od_ma_freqs$Year <- as.character(eu_new_od_ma_freqs$Year)
eu_new_od_ma_freqs$Year <- as.numeric(eu_new_od_ma_freqs$Year)
eu_new_od_ma_freqs <- eu_new_od_ma_freqs[order(-eu_new_od_ma_freqs$Year),] 

head(eu_new_od_ma_freqs)
```

Select all MAs in the EU

format EU market authorizations by year

```{r}
eu_od_ma_freqs <- as.data.frame(table(eu_orphan_drugs$market_authorization_year))
colnames(eu_od_ma_freqs) <- c("Year", "Freq Orphan Drug MAs")
eu_od_ma_freqs

eu_od_ma_freqs$Year <- as.character(eu_od_ma_freqs$Year)
eu_od_ma_freqs$Year <- as.numeric(eu_od_ma_freqs$Year)
eu_od_ma_freqs <- eu_od_ma_freqs[order(-eu_od_ma_freqs$Year),] 
head(eu_od_ma_freqs)

eu_od_ma_freqs <- merge(eu_od_ma_freqs, eu_new_od_ma_freqs, by = "Year")
eu_od_ma_freqs
```

Graph of EU Market Authorizations orphan drugs

```{r}
ggplot(eu_od_ma_freqs, aes(x = Year)) +
  geom_line(aes(y = `Freq Orphan Drug MAs`, color="Freq Orphan Drug MAs")) +
  geom_line(aes(y = `Freq New Orphan Drug MAs`, color="Freq New Orphan Drug MAs")) +
  labs(x = "Year",
       y = "Frequency of Market Authorizations by the EFPIA",
       title="Annual Frequencies of Market Authorizations in the EU", 
       ) +
  scale_color_manual(name = "Legend", values = c("Freq Orphan Drug MAs" = "#5D3A9B", "Freq New Orphan Drug MAs" = "#E66100")) +
  theme_minimal()
```

merge orphan drug and R&D spending tables

```{r}
colnames(efpia_rd_totals) <- c("Year", "total_rd_euros", "total_rd_mil_dollars" , "total_rd_bil_dollars")
eu_rd_mas <- merge(efpia_rd_totals,eu_od_ma_freqs,by="Year")
eu_rd_mas <- clean_names(eu_rd_mas)
eu_rd_mas
```

# PhRMA/ United States

## PhRMA R&D spending by year

<https://phrma.org/-/media/Project/PhRMA/PhRMA-Org/PhRMA-Refresh/Report-PDFs/P-R/PhRMA_membership-survey_2022_final.pdf>

```{r}
phrma_us_rd <- read.csv("Cleaned Data/Area Level R&D Spending/US Domestic R&D and R&D Abroad,* PhRMA Member Companies- 1983–2021 (dollar figures in millions) .csv")
head(phrma_us_rd)
```

## US orphan drug Market Authorizations

<https://www.accessdata.fda.gov/scripts/opdlisting/oopd/> Only approved
products

```{r}
us_orphan_drugs <- read.csv("Original Data/Area Level Orphan Authorizations/US_FDA_OD.csv")

us_orphan_drugs <- clean_names(us_orphan_drugs)
us_orphan_drugs <- us_orphan_drugs[!is.na(us_orphan_drugs$marketing_approval_date), ] 

us_orphan_drugs$marketing_approval_date <- as.Date(us_orphan_drugs$marketing_approval_date, format="%m/%d/%y")

us_orphan_drugs[order(us_orphan_drugs$marketing_approval_date),]


```

Check for duplicated rows Remove one row because duplicated

```{r}
us_orphan_drugs[duplicated(us_orphan_drugs),]
us_orphan_drugs[us_orphan_drugs$generic_name == "tacrolimus" ,]
us_orphan_drugs <- us_orphan_drugs[-1119, ]
us_orphan_drugs[duplicated(us_orphan_drugs),]

```

Number of designations per product

```{r}
us_ma_freq <- us_orphan_drugs %>%
  count(generic_name) %>%
  arrange(desc(n))
colnames(us_ma_freq) <- c("Generic Name", "Market Authorization Frequency")

us_ma_freq
```

How many ODs with more than 1 authorization

```{r}
nrow(us_ma_freq[us_ma_freq$`Market Authorization Frequency` > 1, ])
```

```{r}
stargazer(us_ma_freq[1:19,], title= "US Market Authorization Frequencies for Orphan Drugs with Over 5 Market Authorizations", rownames = FALSE, align=TRUE, summary=FALSE, type = "html", out="Final Figures Formatted/US_OD_Multiple_MAs.htm")

```

Select only new drugs in the US

```{r}
us_orphan_drugs_new <- us_orphan_drugs %>%
  group_by(generic_name) %>%  
  slice_min(order_by = marketing_approval_date, n = 1, with_ties = FALSE) %>%  # Select the entry with the earliest market authorization year
  ungroup() 

us_orphan_drugs_new

```

```{r}
us_orphan_drugs_new$Marketing.year <- year(as.Date(us_orphan_drugs_new$marketing_approval_date)) 
us_new_od_ma_freqs <- as.data.frame(table(us_orphan_drugs_new$Marketing.year))
colnames(us_new_od_ma_freqs) <- c("Year", "Freq New Orphan Drug MAs")
us_new_od_ma_freqs
```

Make sure matches the "Number of Orphan Drug Approvals: 1262" from
<https://www.accessdata.fda.gov/scripts/opdlisting/oopd/listResult.cfm>
on 9/29/2024

```{r}
nrow(us_orphan_drugs[year(as.Date(us_orphan_drugs$marketing_approval_date)), ])

```

Select all MAs in the US format US MA date as year

```{r}
us_orphan_drugs$Marketing.year <- year(as.Date(us_orphan_drugs$marketing_approval_date)) 
us_od_ma_freqs <- as.data.frame(table(us_orphan_drugs$Marketing.year))
colnames(us_od_ma_freqs) <- c("Year", "Freq Orphan Drug MAs")

us_od_ma_freqs$Year <- as.character(us_od_ma_freqs$Year)
us_od_ma_freqs$Year <- as.numeric(us_od_ma_freqs$Year)
us_od_ma_freqs <- us_od_ma_freqs[order(-us_od_ma_freqs$Year),] 

us_od_ma_freqs <- merge(us_od_ma_freqs, us_new_od_ma_freqs, by = "Year")

us_od_ma_freqs
```

Graph of US Market Authorizations orphan drugs

```{r}
ggplot(us_od_ma_freqs, aes(x = Year)) +
  geom_line(aes(y = `Freq Orphan Drug MAs`, color="Freq Orphan Drug MAs")) +
  geom_line(aes(y = `Freq New Orphan Drug MAs`, color="Freq New Orphan Drug MAs")) +
  
  # geom_text(aes(y = `Freq Orphan Drug MAs`, 
  #               label = `Freq Orphan Drug MAs`, 
  #               color = "Freq Orphan Drug MAs"), 
  #           vjust = -2, 
  #           size = 3) +
  # 
  # # Add text labels for the y-values of Freq New Orphan Drug MAs
  # geom_text(aes(y = `Freq New Orphan Drug MAs`, 
  #               label = `Freq New Orphan Drug MAs`, 
  #               color = "Freq New Orphan Drug MAs"), 
  #           vjust = 2, 
  #           size = 3) +
  # 
  
  labs(x = "Year",
       y = "Frequency of Market Authorizations by the FDA",
       title="Annual Frequencies of Market Authorizations in the US", 
       ) +
  scale_color_manual(name = "Legend", values = c('Freq Orphan Drug MAs' = '#5D3A9B', 'Freq New Orphan Drug MAs' = '#E66100')) +
  theme_minimal()
```

```{r}
us_orphan_drugs
```

merge US orphan drug df with PhRMA r&d df

```{r}
phrma_us_rd <- phrma_us_rd[-dim(phrma_us_rd)[1],]
us_rd_mas <- merge(phrma_us_rd,us_od_ma_freqs,by="Year")
us_rd_mas
```

format

```{r}
us_rd_mas$Domestic.R.D. <- parse_number(us_rd_mas$Domestic.R.D.) #dollars are in millions
us_rd_mas <- clean_names(us_rd_mas)
us_rd_mas
```

us dataframe simplified

```{r}
us_rd_mas$annual_domestic_RD_spending_bil_dollars <- us_rd_mas$domestic_r_d /1000 #convert from millions of dollars to billions
us_rd_mas_simplified <- as.data.frame(cbind(us_rd_mas$year, us_rd_mas$annual_domestic_RD_spending_bil_dollars, us_rd_mas$freq_orphan_drug_m_as, us_rd_mas$freq_new_orphan_drug_m_as))
colnames(us_rd_mas_simplified) <- c("year", "annual_domestic_RD_spending_bil_dollars", "freq_orphan_drug_mas", "freq_new_orphan_drug_mas")
us_rd_mas_simplified$area <- rep("US", dim(us_rd_mas_simplified)[1])
us_rd_mas_simplified$lagged_annual_domestic_RD_spending_bil_dollars <- lag(us_rd_mas_simplified$annual_domestic_RD_spending_bil_dollars)
us_rd_mas_simplified
```

eu dataframe simplified

```{r}
eu_rd_mas_simplified <- as.data.frame(cbind(eu_rd_mas$year, eu_rd_mas$total_rd_bil_dollars, eu_rd_mas$freq_orphan_drug_m_as, eu_rd_mas$freq_new_orphan_drug_m_as))
colnames(eu_rd_mas_simplified) <- c("year", "annual_domestic_RD_spending_bil_dollars", "freq_orphan_drug_mas", "freq_new_orphan_drug_mas")
eu_rd_mas_simplified$area <- rep("EU", dim(eu_rd_mas_simplified)[1])
eu_rd_mas_simplified$lagged_annual_domestic_RD_spending_bil_dollars <- lag(eu_rd_mas_simplified$annual_domestic_RD_spending_bil_dollars)
eu_rd_mas_simplified
```

# EU GDP Per Capita In Dollars

<https://www.macrotrends.net/global-metrics/countries/EUU/european-union/gdp-per-capita>

```{r}
eu_gdp_percapita_annual <- read.csv("Original Data/GDP/european-union-gdp-per-capita-annual.csv", skip=16)
eu_gdp_percapita_annual$year <- year(as.Date(eu_gdp_percapita_annual$date))
eu_gdp_percapita_annual <- clean_names(eu_gdp_percapita_annual)
colnames(eu_gdp_percapita_annual)[2] <- "gdp_per_capita_usd"
eu_gdp_percapita_annual
```

# EU GDP Annual Growth Rate

<https://www.macrotrends.net/global-metrics/countries/EUU/european-union/gdp-growth-rate>

```{r}
eu_gdp_growth_rate_annual <- read.csv("Original Data/GDP/european-union-gdp-growth-rate.csv", skip=16)

eu_gdp_growth_rate_annual$year <- year(as.Date(eu_gdp_growth_rate_annual$date)) #format year


eu_gdp_growth_rate_annual <- clean_names(eu_gdp_growth_rate_annual)

colnames(eu_gdp_growth_rate_annual)[2] <- "gdp_growth_rate_annual"

eu_gdp_growth_rate_annual$gdp_growth_rate_annual <- eu_gdp_growth_rate_annual$gdp_growth_rate_annual/100 #convert to rate instead of percent

eu_gdp_growth_rate_annual
```

Merge GDP Per Capita, GDP Growth rate into same df

```{r}
eu_rd_mas_simplified <- merge(eu_rd_mas_simplified, eu_gdp_percapita_annual[,c("year", "gdp_per_capita_usd")], by = "year", all.x = TRUE)
eu_rd_mas_simplified <- merge(eu_rd_mas_simplified, eu_gdp_growth_rate_annual[,c("year", "gdp_growth_rate_annual")], by = "year", all.x = TRUE)
eu_rd_mas_simplified
```

# US GDP Per Capita In Dollars

<https://www.macrotrends.net/global-metrics/countries/USA/united-states/gdp-per-capita>

```{r}
us_gdp_percapita_annual <- read.csv("Original Data/GDP/united-states-gdp-per-capita-annual.csv", skip=16)
us_gdp_percapita_annual$year <- year(as.Date(us_gdp_percapita_annual$date))

us_gdp_percapita_annual <- clean_names(us_gdp_percapita_annual)

colnames(us_gdp_percapita_annual)[2] <- "gdp_per_capita_usd"
us_gdp_percapita_annual

```

# US GDP Annual Growth Rate

<https://www.macrotrends.net/global-metrics/countries/USA/united-states/gdp-growth-rate>

```{r}
us_gdp_growth_rate_annual <- read.csv("Original Data/GDP/united-states-gdp-growth-rate.csv", skip=16)
us_gdp_growth_rate_annual$year <- year(as.Date(us_gdp_growth_rate_annual$date))

us_gdp_growth_rate_annual <- clean_names(us_gdp_growth_rate_annual)

colnames(us_gdp_growth_rate_annual)[2] <- "gdp_growth_rate_annual"
us_gdp_growth_rate_annual$gdp_growth_rate_annual <- us_gdp_growth_rate_annual$gdp_growth_rate_annual/100 #convert to rate instead of percent

us_gdp_growth_rate_annual

```

Merge GDP Per Capita, GDP Growth rate into same df

```{r}
us_rd_mas_simplified <- merge(us_rd_mas_simplified, us_gdp_percapita_annual[,c("year", "gdp_per_capita_usd")], by = "year", all.x = TRUE)
us_rd_mas_simplified <- merge(us_rd_mas_simplified, us_gdp_growth_rate_annual[,c("year", "gdp_growth_rate_annual")], by = "year", all.x = TRUE)
us_rd_mas_simplified
```

```{r}
combined_rd_ma_df <- rbind(eu_rd_mas_simplified, us_rd_mas_simplified)
combined_rd_ma_df <- combined_rd_ma_df %>%
  filter(year > 2003)
combined_rd_ma_df
```

```{r}
colorpallate <- c("#D81B60", "#1E88E5")
ggplot(combined_rd_ma_df, aes(x = annual_domestic_RD_spending_bil_dollars, y = freq_orphan_drug_mas, color = area)) +
  geom_point() +
  labs(x = "Annual Domestic RD Spending (Billions USD)",
       y = "Number of Orphan Drug Market Authorizations",
       color = "Area",
       title=str_wrap("Orphan Drug Market Authorizations by Annual Domestic R&D Spending in the EU and US", 60)) +
  scale_color_manual(values = colorpallate)+
  theme_minimal()
```

```{r}
ggplot(combined_rd_ma_df, aes(x = year , y = freq_orphan_drug_mas, color = area)) +
  geom_line() +
  labs(x = "Year",
       y = "Number of Orphan Drug Market Authorizations",
       color = "Area",
       title="Orphan Drug Market Authorizations by Year in the EU and US") +
    theme_minimal() +
    scale_color_manual(values = colorpallate)
```

```{r}
ggplot(combined_rd_ma_df, aes(x = year , y = freq_new_orphan_drug_mas, color = area)) +
  geom_line() +
  labs(x = "Year",
       y = "Number of Market Authorizations for New Orphan Drugs",
       color = "Area",
       title="Market Authorizations for New Orphan Drugs by Year in the EU and US") +
    theme_minimal() +
    scale_color_manual(values = colorpallate)
```

```{r}
ggplot(combined_rd_ma_df, aes(x = year, y = annual_domestic_RD_spending_bil_dollars, color = area)) +
  geom_line() +
  labs(x = "Year",
       y = "Annual Domestic R&D Spending (Billions USD)",
       color = "Area",
       title="Annual Domestic R&D Spending by Year in the EU and US") +
  scale_color_manual(values = colorpallate)+
  theme_minimal()
```

```{r}
ggplot(combined_rd_ma_df, aes(x = year, y = gdp_per_capita_usd, color = area)) +
  geom_line() +
  labs(x = "Year",
       y = "GDP Per Capita on last day of Year (USD)",
       color = "Area",
       title="GDP Per Capita on last day of Year in the EU and US") +
  scale_color_manual(values = colorpallate)+
  theme_minimal()
```

```{r}
ggplot(combined_rd_ma_df, aes(x = year, y = gdp_growth_rate_annual *100, color = area)) +
  geom_line() +
  labs(x = "Year",
       y = "GDP Annual Growth Rate (%)",
       color = "Area",
       title="GDP Annual Growth Rate over Time In EU and US") +
  scale_color_manual(values = colorpallate)+
  theme_minimal()
```

```{r}
plot(combined_rd_ma_df[,-4], col=as.factor(combined_rd_ma_df[,4]))
```

```{r}
subset(combined_rd_ma_df, area=="US")
```

Orphan Drug Frequencies by Year by Area

```{r}
hist(subset(combined_rd_ma_df, area=="US")$freq_orphan_drug_mas, main="US Frequencies of Annual Orphan Drug Market Authorizations", xlab="Number of Annual US Orphan Drug Market Authorizations")
hist(subset(combined_rd_ma_df, area=="EU")$freq_orphan_drug_mas, main="EU Frequencies of Annual Orphan Drug Market Authorizations",xlab="Number of Annual EU Orphan Drug Market Authorizations")
```

New Orphan Drugs Frequencies by Year by Area

```{r}
hist(subset(combined_rd_ma_df, area=="US")$freq_new_orphan_drug_mas, main="US Frequencies of Annual New Orphan Drug Market Authorizations", xlab="Number of Annual US Orphan Drug Market Authorizations")
hist(subset(combined_rd_ma_df, area=="EU")$freq_new_orphan_drug_mas, main="EU Frequencies of Annual New Orphan Drug Market Authorizations",xlab="Number of Annual EU Orphan Drug Market Authorizations")
```

# Making a years after 2004 variable

```{r}
combined_rd_ma_df$yearsafter2004 <- combined_rd_ma_df$year - 2004
combined_rd_ma_df
```

# R&D Models

## Model of R&D Spending by area

```{r}
mod.rd.1 <- lm(annual_domestic_RD_spending_bil_dollars ~  as.factor(area) , data=combined_rd_ma_df)
summary(mod.rd.1)

residuals <- resid(mod.rd.1)
plot(residuals)

acf(residuals)
pacf(residuals)

mod.rd.1.vcov <- vcovPL(mod.rd.1, cluster = ~ as.factor(combined_rd_ma_df$area), lag="NW1994") 

mod.rd.nw.1 <- coeftest(mod.rd.1, vcov = mod.rd.1.vcov) 
mod.rd.nw.1

```

## Model of R&D spending by country and year without interaction

```{r}
mod.rd.2 <- lm(annual_domestic_RD_spending_bil_dollars ~  as.factor(area) + yearsafter2004, data=combined_rd_ma_df)
summary(mod.rd.2)

residuals <- resid(mod.rd.2)
plot(residuals)

acf(residuals)
pacf(residuals)

mod.rd.2.vcov <- vcovPL(mod.rd.2, cluster = ~ as.factor(combined_rd_ma_df$area), lag="NW1994") 

mod.rd.nw.2 <- coeftest(mod.rd.2, vcov = mod.rd.2.vcov) 
mod.rd.nw.2



```

## Model of R&D spending by country and year with interaction

```{r}
mod.rd.3 <- lm(annual_domestic_RD_spending_bil_dollars ~  as.factor(area) * yearsafter2004, data=combined_rd_ma_df)
summary(mod.rd.3)

residuals <- resid(mod.rd.3)
plot(residuals)

acf(residuals)
pacf(residuals)

mod.rd.3.vcov <- vcovPL(mod.rd.3, cluster = ~ as.factor(combined_rd_ma_df$area), lag="NW1994") 

mod.rd.nw.3 <- coeftest(mod.rd.3, vcov = mod.rd.3.vcov) 
mod.rd.nw.3

```

## Model of R&D spending by area and year adjusted for gdp growth rate

```{r}
mod.rd.4 <- lm(annual_domestic_RD_spending_bil_dollars ~  as.factor(area) + gdp_growth_rate_annual + yearsafter2004 + yearsafter2004:as.factor(area) , data=combined_rd_ma_df)
summary(mod.rd.4)

residuals <- resid(mod.rd.4)
plot(residuals)

acf(residuals)
pacf(residuals)

mod.rd.4.vcov <- vcovPL(mod.rd.4, cluster = ~ as.factor(combined_rd_ma_df$area), lag="NW1994")  
mod.rd.nw.4 <- coeftest(mod.rd.4, vcov = mod.rd.4.vcov) # vcov = NeweyWest(mod.rd.4, lag = 1, prewhite = FALSE))
mod.rd.nw.4

```

<!-- # ```{r} -->

<!-- # mod.rd.nw.1 <- gls(model=annual_domestic_RD_spending_dollars ~  as.factor(area) + gdp_per_capita_usd +yearsafter2004 + yearsafter2004*as.factor(area), data=combined_rd_od_df, corARMA(p = 0, q = 1)) -->

<!-- # summary(mod.rd.nw.1) -->

<!-- # ``` -->

<!-- ```{r} -->

<!-- tidy_mod.rd.nw.1 <- tidy(mod.rd.nw.1) -->

<!-- tidy_mod.rd.nw.1 <- tidy_mod.rd.nw.1 %>% mutate(Model = "Model 1") -->

<!-- combined_models <- bind_rows(tidy_mod.rd.nw.1) -->

<!-- formatted_table <- combined_models %>% -->

<!--   select(Model, term, estimate, std.error, statistic, p.value) %>% -->

<!--   pivot_wider(names_from = Model, values_from = c(estimate, std.error, statistic, p.value)) -->

<!-- ``` -->

<!-- ```{r} -->

<!-- formatted_table -->

<!-- ``` -->

```{r}
stargazer(mod.rd.1, mod.rd.2, mod.rd.3, mod.rd.4, title="Models of R&D Spending (Not Adjusted for Error Structure)", align=TRUE, type = "html", out="Final Models Formatted/rdmodelsNonAdjusted.htm",
          covariate.labels = c("US", "Annual GDP Growth Rate", "Years After 2004", "US:Years After 2004", "Intercept"),
          dep.var.labels=c("Annual R&D Spending (Billion USD)"))
```

```{r}
stargazer(mod.rd.nw.1, mod.rd.nw.2, mod.rd.nw.3, mod.rd.nw.4, title="Newey-West Adjusted Models of R&D Spending", align=TRUE, type = "html", out="Final Models Formatted/rdmodels.htm",
          covariate.labels = c("US", "Annual GDP Growth Rate", "Years After 2004", "US:Years After 2004", "Intercept"),
          ci = TRUE,
          #column.labels = c( "Model 1", "Model 2", "Model 3", "Model 4", "Model 5"),
          digits = 2,
          dep.var.labels=c("Annual R&D Spending (Billion USD)"))
```

# Number of Orphan Drug Market Authorization Models

## Freq of orphan drug mas by Domestic R&D, Area, and interaction, year

```{r}
mod.od.1 <- lm(freq_orphan_drug_mas ~ annual_domestic_RD_spending_bil_dollars + as.factor(area)*yearsafter2004 + as.factor(area)*annual_domestic_RD_spending_bil_dollars  , data=combined_rd_ma_df)
summary(mod.od.1)

residuals <- resid(mod.od.1)
plot(residuals)

acf(residuals)
pacf(residuals)
 
mod.od.1.vcov <- vcovPL(mod.od.1, cluster = ~ as.factor(combined_rd_ma_df$area), lag="NW1994")  
mod.od.nw.1 <- coeftest(mod.od.1, vcov = mod.od.1.vcov) 
mod.od.nw.1

confint(mod.od.nw.1)

```

<!-- plm package -->

<!-- ```{r} -->

<!-- mod.rd.od <- plm(freq_orphan_drugs ~ as.numeric(annual_domestic_RD_spending_dollars) + as.factor(area) + as.factor(area)*as.numeric(annual_domestic_RD_spending_dollars), data=combined_rd_od_df, model="pooling") -->

<!-- #summary(mod.rd.od, vcov = vcovNW) -->

<!-- coeftest(mod.rd.od, vcov.=vcovNW) -->

<!-- ``` -->

-   take out interaction and see what the indicator is
-   then put in year as a factor

## Freq of orphan drug mas by Domestic R&D, Area, and interaction and GDP growth rate

```{r}
mod.od.2 <- lm(freq_orphan_drug_mas ~ annual_domestic_RD_spending_bil_dollars + as.factor(area)*yearsafter2004 + gdp_growth_rate_annual + as.factor(area)*annual_domestic_RD_spending_bil_dollars , data=combined_rd_ma_df)
summary(mod.od.2)

residuals <- resid(mod.od.2)
plot(residuals)

acf(residuals)
pacf(residuals)

mod.od.2.vcov <- vcovPL(mod.od.2, cluster = ~ as.factor(combined_rd_ma_df$area), lag="NW1994")  
mod.od.nw.2 <- coeftest(mod.od.2, vcov = mod.od.2.vcov)
mod.od.nw.2
confint(mod.od.nw.2)

```

With 95% confidence, after adjusting for annual R&D spending and annual
GDP growth rate, a one-year increase is associated with between a
further 1.19 and 7 million dollar increase in the number of
market-authorized orphan drugs compared to the EU

With 95% confidence, after adjusting for annual R&D spending and annual
GDP growth rate, a one-year increase is associated with between a
further 1.19 and 7 million dollar increase in the number of
market-authorized orphan drugs compared to the EU

## Freq of orphan drug mas by Domestic R&D in the US with GDP growth, no area/ spending interaction

```{r}
mod.od.3 <- lm(freq_orphan_drug_mas ~ annual_domestic_RD_spending_bil_dollars + as.factor(area) * yearsafter2004 + gdp_growth_rate_annual  , data=combined_rd_ma_df)
summary(mod.od.3)

residuals <- resid(mod.od.3)
plot(residuals)

acf(residuals)
pacf(residuals)

mod.od.3.vcov <- vcovPL(mod.od.3, cluster = ~ as.factor(combined_rd_ma_df$area), lag="NW1994")  
mod.od.nw.3 <-coeftest(mod.od.3, vcov = mod.od.3.vcov)
mod.od.nw.3

confint(mod.od.nw.3)
```

## Freq of orphan drug mas by Domestic R&D in the US no area/ spending interaction or gdp growth rate

```{r}
mod.od.4 <- lm(freq_orphan_drug_mas ~ annual_domestic_RD_spending_bil_dollars + as.factor(area) * yearsafter2004  , data=combined_rd_ma_df)
summary(mod.od.4)

residuals <- resid(mod.od.4)
plot(residuals)

acf(residuals)
pacf(residuals)

mod.od.4.vcov <- vcovPL(mod.od.4, cluster = ~ as.factor(combined_rd_ma_df$area), lag="NW1994")  
mod.od.nw.4 <-coeftest(mod.od.4, vcov = mod.od.4.vcov)
mod.od.nw.4

confint(mod.od.nw.4)
```

## Freq of orphan drug mas by US over time - no interaction

```{r}
mod.od.5 <- lm(freq_orphan_drug_mas ~ as.factor(area) + yearsafter2004  , data=combined_rd_ma_df)

residuals <- resid(mod.od.5)
plot(residuals)

acf(residuals)
pacf(residuals)

mod.od.5.vcov <- vcovPL(mod.od.5, cluster = ~ as.factor(combined_rd_ma_df$area), lag="NW1994")  
mod.od.nw.5 <-coeftest(mod.od.5, vcov = mod.od.5.vcov)
mod.od.nw.5

confint(mod.od.nw.5)
```

## Freq of orphan drug mas by US over time - interaction

```{r}
mod.od.6 <- lm(freq_orphan_drug_mas ~ as.factor(area) * yearsafter2004  , data=combined_rd_ma_df)

residuals <- resid(mod.od.6)
plot(residuals)

acf(residuals)
pacf(residuals)

mod.od.6.vcov <- vcovPL(mod.od.6, cluster = ~ as.factor(combined_rd_ma_df$area), lag="NW1994")  
mod.od.nw.6 <-coeftest(mod.od.6, vcov = mod.od.6.vcov)
mod.od.nw.6

confint(mod.od.nw.6)
```

```{r}
aic_values <- round(c(AIC(mod.od.5), AIC(mod.od.6), AIC(mod.od.4), AIC(mod.od.1), AIC(mod.od.3), AIC(mod.od.2) ))

stargazer(mod.od.5, mod.od.6, mod.od.4, mod.od.1, mod.od.3, mod.od.2, title="Models of Orphan Drug Market Authorizations (Not Adjusted for Error Structure)", align=TRUE, type = "html", out="Final Models Formatted/orphandrugMAmodelsNonAdjusted.htm",
          covariate.labels = c("Annual Domestic R&D Spending (Billions USD)", "US", "Years After 2004",  "Annual GDP Growth Rate", "Years After 2004:US", "Annual Domestic R&D Spending (Billions USD):US" , "Intercept"),
          dep.var.labels=c("Annual Number of Market Authorized Orphan Drugs"),
          add.lines = list(c("AIC", aic_values))
          )
```

```{r}
#creating shorter names because stargazer has a length limit for the input of all models

m1 <- mod.od.nw.1
m2 <- mod.od.nw.2
m3 <- mod.od.nw.3
m4 <- mod.od.nw.4
m5 <- mod.od.nw.5 
m6 <- mod.od.nw.6

aic_values <- round(c(AIC(m5), AIC(m6), AIC(m4), AIC(m1), AIC(m3), AIC(m2) ))


stargazer(m5,m6,m4,m1,m3,m2, title="Newey-West Adjusted Models of Orphan Drug Market Authorizations", align=TRUE, type = "html", out="Final Models Formatted/orphandrugMAmodels.htm",
          covariate.labels = c("Annual Domestic R&D Spending (Billions USD)", "US", "Years After 2004",  "Annual GDP Growth Rate", "Years After 2004:US", "Annual Domestic R&D Spending (Billions USD):US" , "Intercept"),
          dep.var.labels=c("Annual Number of Market Authorized Orphan Drugs"),
          add.lines = list(c("AIC", aic_values))
          )
```

# Number of Market Authorizations for NEW Orphan Drugs Models

## Freq of mas for NEW orphan drugs by Domestic R&D, Area, and interaction, year

```{r}
mod.new.od.1 <- lm(freq_new_orphan_drug_mas ~ annual_domestic_RD_spending_bil_dollars + as.factor(area)*yearsafter2004 + as.factor(area)*annual_domestic_RD_spending_bil_dollars  , data=combined_rd_ma_df)
summary(mod.new.od.1)

residuals <- resid(mod.new.od.1)
plot(residuals)

acf(residuals)
pacf(residuals)

mod.new.od.1.vcov <- vcovPL(mod.new.od.1, cluster = ~ as.factor(combined_rd_ma_df$area), lag="NW1994")  
mod.new.od.nw.1 <-coeftest(mod.new.od.1, vcov = mod.new.od.1.vcov)
mod.new.od.nw.1

confint(mod.new.od.nw.1)

```

<!-- plm package -->

<!-- ```{r} -->

<!-- mod.rd.od <- plm(freq_orphan_drugs ~ as.numeric(annual_domestic_RD_spending_dollars) + as.factor(area) + as.factor(area)*as.numeric(annual_domestic_RD_spending_dollars), data=combined_rd_od_df, model="pooling") -->

<!-- #summary(mod.rd.od, vcov = vcovNW) -->

<!-- coeftest(mod.rd.od, vcov.=vcovNW) -->

<!-- ``` -->

-   take out interaction and see what the indicator is
-   then put in year as a factor

## Freq of mas for NEW orphan drugs by Domestic R&D, Area, and interaction and GDP growth rate

```{r}
mod.new.od.2 <- lm(freq_new_orphan_drug_mas ~ annual_domestic_RD_spending_bil_dollars + as.factor(area)*yearsafter2004 + gdp_growth_rate_annual + as.factor(area)*annual_domestic_RD_spending_bil_dollars , data=combined_rd_ma_df)
summary(mod.new.od.2)

residuals <- resid(mod.new.od.2)
plot(residuals)

acf(residuals)
pacf(residuals)

mod.new.od.2.vcov <- vcovPL(mod.new.od.2, cluster = ~ as.factor(combined_rd_ma_df$area), lag="NW1994")  
mod.new.od.nw.2 <-coeftest(mod.new.od.2, vcov = mod.new.od.2.vcov)
mod.new.od.nw.2

confint(mod.new.od.nw.2)

```

With 95% confidence, after adjusting for annual R&D spending and annual
GDP growth rate, a one-year increase is associated with between a
further 1.19 and 7 million dollar increase in the number of
market-authorized orphan drugs compared to the EU

With 95% confidence, after adjusting for annual R&D spending and annual
GDP growth rate, a one-year increase is associated with between a
further 1.19 and 7 million dollar increase in the number of
market-authorized orphan drugs compared to the EU

## Freq of mas for NEW orphan drug by Domestic R&D in the US with GDP growth, no area/ spending interaction

```{r}
mod.new.od.3 <- lm(freq_new_orphan_drug_mas ~ annual_domestic_RD_spending_bil_dollars + as.factor(area) * yearsafter2004 + gdp_growth_rate_annual  , data=combined_rd_ma_df)
summary(mod.new.od.3)

residuals <- resid(mod.new.od.3)
plot(residuals)

acf(residuals)
pacf(residuals)

mod.new.od.3.vcov <- vcovPL(mod.new.od.3, cluster = ~ as.factor(combined_rd_ma_df$area), lag="NW1994")  
mod.new.od.nw.3 <-coeftest(mod.new.od.3, vcov = mod.new.od.3.vcov)
mod.new.od.nw.3

confint(mod.new.od.nw.3)
```

## Freq of orphan drug mas by Domestic R&D in the US no area/ spending interaction or gdp growth rate

```{r}
mod.new.od.4 <- lm(freq_new_orphan_drug_mas ~ annual_domestic_RD_spending_bil_dollars + as.factor(area) * yearsafter2004  , data=combined_rd_ma_df)
summary(mod.new.od.4)

residuals <- resid(mod.new.od.4)
plot(residuals)

acf(residuals)
pacf(residuals)


mod.new.od.4.vcov <- vcovPL(mod.new.od.4, cluster = ~ as.factor(combined_rd_ma_df$area), lag="NW1994")  
mod.new.od.nw.4 <-coeftest(mod.new.od.4, vcov = mod.new.od.4.vcov)
mod.new.od.nw.4

confint(mod.new.od.nw.4)
```

## Freq of mas for NEW orphan drugs by US over time - no interaction

```{r}
mod.new.od.5 <- lm(freq_new_orphan_drug_mas ~ as.factor(area) + yearsafter2004  , data=combined_rd_ma_df)

residuals <- resid(mod.new.od.5)
plot(residuals)

acf(residuals)
pacf(residuals)

mod.new.od.5.vcov <- vcovPL(mod.new.od.5, cluster = ~ as.factor(combined_rd_ma_df$area), lag="NW1994")  
mod.new.od.nw.5 <-coeftest(mod.new.od.5, vcov = mod.new.od.5.vcov)
mod.new.od.nw.5

confint(mod.new.od.nw.5)
```

## Freq of NEW orphan drugs mas by US over time - interaction

```{r}
mod.new.od.6 <- lm(freq_new_orphan_drug_mas ~ as.factor(area) * yearsafter2004  , data=combined_rd_ma_df)

residuals <- resid(mod.new.od.6)
plot(residuals)

acf(residuals)
pacf(residuals)


mod.new.od.6.vcov <- vcovPL(mod.new.od.6, cluster = ~ as.factor(combined_rd_ma_df$area), lag="NW1994")  
mod.new.od.nw.6 <-coeftest(mod.new.od.6, vcov = mod.new.od.6.vcov)
mod.new.od.nw.6


confint(mod.new.od.nw.6)
```

```{r}
#creating shorter names because stargazer has a length limit for the input of all models

m1 <- mod.new.od.1
m2 <- mod.new.od.2
m3 <- mod.new.od.3
m4 <- mod.new.od.4
m5 <- mod.new.od.5 
m6 <- mod.new.od.6

aic_values <- round(c(AIC(m5), AIC(m6), AIC(m4), AIC(m1), AIC(m3), AIC(m2) ))


stargazer(m5, m6, m4, m1, m3, m2, title="Models of Market Authorizations for NEW Orphan Drugs (Not adjusted for Error Structure)", align=TRUE, type = "html", out="Final Models Formatted/NEWorphandrugMAmodelsNonAdjusted.htm",
          covariate.labels = c("Annual Domestic R&D Spending (Billions USD)", "US", "Years After 2004",  "Annual GDP Growth Rate", "Years After 2004:US", "Annual Domestic R&D Spending (Billions USD):US" , "Intercept"),
          dep.var.labels=c("Annual Number of Market Authorizations for New Orphan Drugs"),
          add.lines = list(c("AIC", aic_values))

          )
```

```{r}
#creating shorter names because stargazer has a length limit for the input of all models

m1 <- mod.new.od.nw.1
m2 <- mod.new.od.nw.2
m3 <- mod.new.od.nw.3
m4 <- mod.new.od.nw.4
m5 <- mod.new.od.nw.5 
m6 <- mod.new.od.nw.6

aic_values <- round(c(AIC(m5), AIC(m6), AIC(m4), AIC(m1), AIC(m3), AIC(m2) ))


stargazer(m5, m6, m4, m1, m3, m2, title="Newey-West Adjusted Models of Market Authorizations for NEW Orphan Drugs", align=TRUE, type = "html", out="Final Models Formatted/NEWorphandrugMAmodels.htm",
          covariate.labels = c("Annual Domestic R&D Spending (Billions USD)", "US", "Years After 2004",  "Annual GDP Growth Rate", "Years After 2004:US", "Annual Domestic R&D Spending (Billions USD):US" , "Intercept"),
          dep.var.labels=c("Annual Number of Market Authorizations for New Orphan Drugs"),
          add.lines = list(c("AIC", aic_values)))
```

-   interaction btw area and r&d

-   look in literature and see if policy chances list policy change by
    year

-   intervention shock that stays - indicator 1 where policy is 0 where
    its not

-   line plot instead of points

-   subset on same time frame - find eu data if possible but if not
    available then subset us

-   add graphs and stuff to lit review

-   add tables to

-   change gdp to % change in gdp - quarterly Federal reserve FRED

```{r}
combined_rd_ma_df
```

Overall mean of Domestic R&D spending

```{r}
mean(combined_rd_ma_df$annual_domestic_RD_spending_bil_dollars)
```

Overall mean of US Domestic R&D spending

```{r}
mean(subset(combined_rd_ma_df ,area=="US")$annual_domestic_RD_spending_bil_dollars)
```

Overall mean of annual EU Domestic R&D spending

```{r}
mean(subset(combined_rd_ma_df ,area=="EU")$annual_domestic_RD_spending_bil_dollars)
```

Overall mean of annual number of US Market Authroized Orphan Drugs

```{r}
mean(subset(combined_rd_ma_df ,area=="US")$freq_orphan_drug_mas)
mean(subset(combined_rd_ma_df ,area=="US")$freq_new_orphan_drug_mas)
```

Overall mean of annual number of EU Market Authroized Orphan Drugs

```{r}
mean(subset(combined_rd_ma_df ,area=="EU")$freq_orphan_drug_mas)
mean(subset(combined_rd_ma_df ,area=="EU")$freq_new_orphan_drug_mas)

```

# Poisson Models

# Poisson Models for Number of Orphan Drug Market Authorization Models

## Poisson model of Freq of orphan drug mas by Domestic R&D, Area, and interaction, year

```{r}
mod.pois.od.1 <- glm(freq_orphan_drug_mas ~ annual_domestic_RD_spending_bil_dollars + as.factor(area)*yearsafter2004 + as.factor(area)*annual_domestic_RD_spending_bil_dollars  , data=combined_rd_ma_df, family=poisson(link="log")) 
summary(mod.pois.od.1)

residuals <- resid(mod.pois.od.1)
plot(residuals)

acf(residuals)
pacf(residuals)
 
mod.pois.od.1.vcov <- vcovPL(mod.pois.od.1, cluster = ~ as.factor(combined_rd_ma_df$area), lag="NW1994")  
mod.pois.od.nw.1 <- coeftest(mod.pois.od.1, vcov = mod.pois.od.1.vcov)
mod.pois.od.nw.1

confint(mod.pois.od.nw.1)

```

<!-- plm package -->

<!-- ```{r} -->

<!-- mod.rd.od <- plm(freq_orphan_drugs ~ as.numeric(annual_domestic_RD_spending_dollars) + as.factor(area) + as.factor(area)*as.numeric(annual_domestic_RD_spending_dollars), data=combined_rd_od_df, model="pooling") -->

<!-- #summary(mod.rd.od, vcov = vcovNW) -->

<!-- coeftest(mod.rd.od, vcov.=vcovNW) -->

<!-- ``` -->

-   take out interaction and see what the indicator is
-   then put in year as a factor

## Poisson model of Freq of orphan drug mas by Domestic R&D, Area, and interaction and GDP growth rate

```{r}
mod.pois.od.2 <- glm(freq_orphan_drug_mas ~ annual_domestic_RD_spending_bil_dollars + as.factor(area)*yearsafter2004 + gdp_growth_rate_annual + as.factor(area)*annual_domestic_RD_spending_bil_dollars , data=combined_rd_ma_df, family=poisson(link="log")) 
summary(mod.pois.od.2)

residuals <- resid(mod.pois.od.2)
plot(residuals)

acf(residuals)
pacf(residuals)

 
mod.pois.od.2.vcov <- vcovPL(mod.pois.od.2, cluster = ~ as.factor(combined_rd_ma_df$area), lag="NW1994")  
mod.pois.od.nw.2 <- coeftest(mod.pois.od.2, vcov = mod.pois.od.2.vcov)
mod.pois.od.nw.2

confint(mod.pois.od.nw.2)

```

## Poisson model of Freq of orphan drug mas by Domestic R&D in the US with GDP growth, no area/ spending interaction

```{r}
mod.pois.od.3 <- glm(freq_orphan_drug_mas ~ annual_domestic_RD_spending_bil_dollars + as.factor(area) * yearsafter2004 + gdp_growth_rate_annual  , data=combined_rd_ma_df, family=poisson(link="log")) 
summary(mod.pois.od.3)

residuals <- resid(mod.pois.od.3)
plot(residuals)

acf(residuals)
pacf(residuals)

mod.pois.od.3.vcov <- vcovPL(mod.pois.od.3, cluster = ~ as.factor(combined_rd_ma_df$area), lag="NW1994")  
mod.pois.od.nw.3 <- coeftest(mod.pois.od.3, vcov = mod.pois.od.3.vcov)
mod.pois.od.nw.3

confint(mod.pois.od.nw.3)
```

## Poisson Model of Freq of orphan drug mas by Domestic R&D in the US no area/ spending interaction or gdp growth rate

```{r}
mod.pois.od.4 <- glm(freq_orphan_drug_mas ~ annual_domestic_RD_spending_bil_dollars + as.factor(area) * yearsafter2004  , data=combined_rd_ma_df , family=poisson(link="log"))
summary(mod.pois.od.4)

residuals <- resid(mod.pois.od.4)
plot(residuals)

acf(residuals)
pacf(residuals)

mod.pois.od.4.vcov <- vcovPL(mod.pois.od.4, cluster = ~ as.factor(combined_rd_ma_df$area), lag="NW1994")  
mod.pois.od.nw.4 <- coeftest(mod.pois.od.4, vcov = mod.pois.od.4.vcov)
mod.pois.od.nw.4

confint(mod.pois.od.nw.4)
```

## Poisson Model for Freq of orphan drug mas by US over time - no interaction

```{r}
mod.pois.od.5 <- glm(freq_orphan_drug_mas ~ as.factor(area) + yearsafter2004  , data=combined_rd_ma_df, family=poisson(link="log"))

residuals <- resid(mod.pois.od.5)
plot(residuals)

acf(residuals)
pacf(residuals)

mod.pois.od.5.vcov <- vcovPL(mod.pois.od.5, cluster = ~ as.factor(combined_rd_ma_df$area), lag="NW1994")  
mod.pois.od.nw.5 <- coeftest(mod.pois.od.5, vcov = mod.pois.od.5.vcov)
mod.pois.od.nw.5

confint(mod.pois.od.nw.5)
```

## Poisson Model For Freq of orphan drug mas by US over time - interaction

```{r}
mod.pois.od.6 <- glm(freq_orphan_drug_mas ~ as.factor(area) * yearsafter2004  , data=combined_rd_ma_df, family=poisson(link="log"))

residuals <- resid(mod.pois.od.6)
plot(residuals)

acf(residuals)
pacf(residuals)

mod.pois.od.6.vcov <- vcovPL(mod.pois.od.6, cluster = ~ as.factor(combined_rd_ma_df$area), lag="NW1994")  
mod.pois.od.nw.6 <- coeftest(mod.pois.od.6, vcov = mod.pois.od.6.vcov)
mod.pois.od.nw.6

confint(mod.pois.od.nw.6)
```

```{r}

#creating shorter names because stargazer has a length limit for the input of all models

m1 <- mod.pois.od.1
m2 <- mod.pois.od.2
m3 <- mod.pois.od.3
m4 <- mod.pois.od.4
m5 <- mod.pois.od.5 
m6 <- mod.pois.od.6

stargazer(m5, m6, m4, m1, m3, m2, title="Poisson Models of Orphan Drug Market Authorizations (Not adjusted for Error Structure)", align=TRUE, type = "html", out="Final Models Formatted/PoissonorphandrugMAmodelsNonAdjusted.htm",
          covariate.labels = c("Annual Domestic R&D Spending (Billions USD)", "US", "Years After 2004",  "Annual GDP Growth Rate", "Years After 2004:US", "Annual Domestic R&D Spending (Billions USD):US" , "Intercept"),
          dep.var.labels=c("Annual Number of Market Authorized Orphan Drugs")
          )
```

```{r}
#creating shorter names because stargazer has a length limit for the input of all models

m1 <- mod.pois.od.nw.1
m2 <- mod.pois.od.nw.2
m3 <- mod.pois.od.nw.3
m4 <- mod.pois.od.nw.4
m5 <- mod.pois.od.nw.5 
m6 <- mod.pois.od.nw.6

aic_values <- round(c(AIC(m5), AIC(m6), AIC(m4), AIC(m1), AIC(m3), AIC(m2) ))

stargazer(m5,m6,m4,m1,m3,m2, title="Newey-West Adjusted Poisson Models of Orphan Drug Market Authorizations", align=TRUE, type = "html", out="Final Models Formatted/PoissonorphandrugMAmodels.htm",
          covariate.labels = c("Annual Domestic R&D Spending (Billions USD)", "US", "Years After 2004",  "Annual GDP Growth Rate", "Years After 2004:US", "Annual Domestic R&D Spending (Billions USD):US" , "Intercept"),
          dep.var.labels=c("Annual Number of Market Authorized Orphan Drugs"),
          add.lines = list(c("AIC", aic_values))

          )
```

# Poisson Models for Number of Market Authorizations for NEW Orphan Drugs

## Poisson model of Freq of mas for NEW Orphan Drugs by Domestic R&D, Area, and interaction, year

```{r}
mod.pois.new.od.1 <- glm(freq_new_orphan_drug_mas ~ annual_domestic_RD_spending_bil_dollars + as.factor(area)*yearsafter2004 + as.factor(area)*annual_domestic_RD_spending_bil_dollars  , data=combined_rd_ma_df, family=poisson(link="log")) 
summary(mod.pois.new.od.1)

residuals <- resid(mod.pois.new.od.1)
plot(residuals)

acf(residuals)
pacf(residuals)

mod.pois.new.od.1.vcov <- vcovPL(mod.pois.new.od.1, cluster = ~ as.factor(combined_rd_ma_df$area), lag="NW1994")  
mod.pois.new.od.nw.1 <- coeftest(mod.pois.new.od.1, vcov = mod.pois.new.od.1.vcov)
mod.pois.new.od.nw.1

confint(mod.pois.new.od.nw.1)
```

## Poisson model of Freq of mas for NEW Orphan Drugs by Domestic R&D, Area, and interaction and GDP growth rate

```{r}
mod.pois.new.od.2 <- glm(freq_new_orphan_drug_mas ~ annual_domestic_RD_spending_bil_dollars + as.factor(area)*yearsafter2004 + gdp_growth_rate_annual + as.factor(area)*annual_domestic_RD_spending_bil_dollars , data=combined_rd_ma_df, family=poisson(link="log")) 
summary(mod.pois.new.od.2)

residuals <- resid(mod.pois.new.od.2)
plot(residuals)

acf(residuals)
pacf(residuals)

mod.pois.new.od.2.vcov <- vcovPL(mod.pois.new.od.2, cluster = ~ as.factor(combined_rd_ma_df$area), lag="NW1994")  
mod.pois.new.od.nw.2 <- coeftest(mod.pois.new.od.2, vcov = mod.pois.new.od.2.vcov)
mod.pois.new.od.nw.2

confint(mod.pois.new.od.nw.2)

```

## Poisson model of Freq of mas for NEW Orphan Drugs by Domestic R&D in the US with GDP growth, no area/ spending interaction

```{r}
mod.pois.new.od.3 <- glm(freq_new_orphan_drug_mas ~ annual_domestic_RD_spending_bil_dollars + as.factor(area) * yearsafter2004 + gdp_growth_rate_annual  , data=combined_rd_ma_df, family=poisson(link="log")) 
summary(mod.pois.new.od.3)

residuals <- resid(mod.pois.new.od.3)
plot(residuals)

acf(residuals)
pacf(residuals)

mod.pois.new.od.3.vcov <- vcovPL(mod.pois.new.od.3, cluster = ~ as.factor(combined_rd_ma_df$area), lag="NW1994")  
mod.pois.new.od.nw.3 <- coeftest(mod.pois.new.od.3, vcov = mod.pois.new.od.3.vcov)
mod.pois.new.od.nw.3


confint(mod.pois.new.od.nw.3)
```

## Poisson Model of Freq of mas for NEW Orphan Drugs by Domestic R&D in the US no area/ spending interaction or gdp growth rate

```{r}
mod.pois.new.od.4 <- glm(freq_new_orphan_drug_mas ~ annual_domestic_RD_spending_bil_dollars + as.factor(area) * yearsafter2004  , data=combined_rd_ma_df , family=poisson(link="log"))
summary(mod.pois.new.od.4)

residuals <- resid(mod.pois.new.od.4)
plot(residuals)

acf(residuals)
pacf(residuals)

mod.pois.new.od.4.vcov <- vcovPL(mod.pois.new.od.4, cluster = ~ as.factor(combined_rd_ma_df$area), lag="NW1994")  
mod.pois.new.od.nw.4 <- coeftest(mod.pois.new.od.4, vcov = mod.pois.new.od.4.vcov)
mod.pois.new.od.nw.4

confint(mod.pois.new.od.nw.4)
```

## Poisson Model for Freq of mas for NEW orphan drugs by US over time - no interaction

```{r}
mod.pois.new.od.5 <- glm(freq_new_orphan_drug_mas ~ as.factor(area) + yearsafter2004  , data=combined_rd_ma_df, family=poisson(link="log"))

residuals <- resid(mod.pois.new.od.5)
plot(residuals)

acf(residuals)
pacf(residuals)

mod.pois.new.od.5.vcov <- vcovPL(mod.pois.new.od.5, cluster = ~ as.factor(combined_rd_ma_df$area), lag="NW1994")  
mod.pois.new.od.nw.5 <- coeftest(mod.pois.new.od.5, vcov = mod.pois.new.od.5.vcov)
mod.pois.new.od.nw.5


confint(mod.pois.new.od.nw.5)
```

## Poisson Model For Freq of mas for NEW Orphan Drugs by US over time - interaction

```{r}
mod.pois.new.od.6 <- glm(freq_new_orphan_drug_mas ~ as.factor(area) * yearsafter2004  , data=combined_rd_ma_df, family=poisson(link="log"))

residuals <- resid(mod.pois.new.od.6)
plot(residuals)

acf(residuals)
pacf(residuals)

mod.pois.new.od.6.vcov <- vcovPL(mod.pois.new.od.6, cluster = ~ as.factor(combined_rd_ma_df$area), lag="NW1994")  
mod.pois.new.od.nw.6 <- coeftest(mod.pois.new.od.6, vcov = mod.pois.new.od.6.vcov)
mod.pois.new.od.nw.6


confint(mod.pois.new.od.nw.6)
```

```{r}

#creating shorter names because stargazer has a length limit for the input of all models

m1 <- mod.pois.new.od.1
m2 <- mod.pois.new.od.2
m3 <- mod.pois.new.od.3
m4 <- mod.pois.new.od.4
m5 <- mod.pois.new.od.5 
m6 <- mod.pois.new.od.6

stargazer(m5, m6, m4, m1, m3, m2, title="Poisson Models of Market Authorizations for NEW Orphan Drugs (Not adjusted for Error Structure)", align=TRUE, type = "html", out="Final Models Formatted/PoissonNEWorphandrugMAmodelsNonAdjusted.htm",
          covariate.labels = c("Annual Domestic R&D Spending (Billions USD)", "US", "Years After 2004",  "Annual GDP Growth Rate", "Years After 2004:US", "Annual Domestic R&D Spending (Billions USD):US" , "Intercept"),
          dep.var.labels=c("Annual Number of Market Authorizations for NEW Orphan Drugs")
          )
```

```{r}
#creating shorter names because stargazer has a length limit for the input of all models

m1 <- mod.pois.new.od.nw.1
m2 <- mod.pois.new.od.nw.2
m3 <- mod.pois.new.od.nw.3
m4 <- mod.pois.new.od.nw.4
m5 <- mod.pois.new.od.nw.5 
m6 <- mod.pois.new.od.nw.6

aic_values <- round(c(AIC(m5), AIC(m6), AIC(m4), AIC(m1), AIC(m3), AIC(m2) ))

stargazer(m5,m6,m4,m1,m3,m2, title="Newey-West Adjusted Poisson Models of Market Authorizations for NEW Orphan Drugs ", align=TRUE, type = "html", out="Final Models Formatted/PoissonNEWorphandrugMAmodels.htm",
          covariate.labels = c("Annual Domestic R&D Spending (Billions USD)", "US", "Years After 2004",  "Annual GDP Growth Rate", "Years After 2004:US", "Annual Domestic R&D Spending (Billions USD):US" , "Intercept"),
          dep.var.labels=c("Annual Number of Market Authorizations for NEW Orphan Drugs"),
          add.lines = list(c("AIC", aic_values))
          )
```



# Stationarity Test

```{r}
kpss.test(combined_rd_ma_df[combined_rd_ma_df$area == "US",]$freq_new_orphan_drug_mas)
kpss.test(combined_rd_ma_df[combined_rd_ma_df$area == "EU",]$freq_new_orphan_drug_mas)

kpss.test(combined_rd_ma_df[combined_rd_ma_df$area == "US",]$freq_orphan_drug_mas)
kpss.test(combined_rd_ma_df[combined_rd_ma_df$area == "EU",]$freq_orphan_drug_mas)

kpss.test(combined_rd_ma_df[combined_rd_ma_df$area == "US",]$annual_domestic_RD_spending_bil_dollars)
kpss.test(combined_rd_ma_df[combined_rd_ma_df$area == "EU",]$annual_domestic_RD_spending_bil_dollars)

```

Low pvalues from Kwiatkowski-Phillips-Schmidt-Shin (KPSS) test for the
null hypothesis that x is level or trend stationary indicate
nonstationarity in the timeseries of all three response variables

Note: To estimate sigma\^2 the Newey-West estimator is used.

<https://www.statisticshowto.com/kpss-test/> "A major disadvantage for
the KPSS test is that it has a high rate of Type I errors (it tends to
reject the null hypothesis too often)."

```{r}
adf.test(combined_rd_ma_df[combined_rd_ma_df$area == "US",]$freq_new_orphan_drug_mas)
adf.test(combined_rd_ma_df[combined_rd_ma_df$area == "EU",]$freq_new_orphan_drug_mas)

adf.test(combined_rd_ma_df[combined_rd_ma_df$area == "US",]$freq_orphan_drug_mas)
adf.test(combined_rd_ma_df[combined_rd_ma_df$area == "EU",]$freq_orphan_drug_mas)

adf.test(combined_rd_ma_df[combined_rd_ma_df$area == "US",]$annual_domestic_RD_spending_bil_dollars)
adf.test(combined_rd_ma_df[combined_rd_ma_df$area == "EU",]$annual_domestic_RD_spending_bil_dollars)

```

Augmented Dickey-Fuller test for the null that x has a unit root.













# ACFs of original series

```{r}
acf(combined_rd_ma_df[combined_rd_ma_df$area == "US",]$freq_new_orphan_drug_mas)
acf(combined_rd_ma_df[combined_rd_ma_df$area == "EU",]$freq_new_orphan_drug_mas)

acf(combined_rd_ma_df[combined_rd_ma_df$area == "US",]$freq_orphan_drug_mas)
acf(combined_rd_ma_df[combined_rd_ma_df$area == "EU",]$freq_orphan_drug_mas)

acf(combined_rd_ma_df[combined_rd_ma_df$area == "US",]$annual_domestic_RD_spending_bil_dollars)
acf(combined_rd_ma_df[combined_rd_ma_df$area == "EU",]$annual_domestic_RD_spending_bil_dollars)
```

# PACFs of original series

```{r}
pacf(combined_rd_ma_df[combined_rd_ma_df$area == "US",]$freq_new_orphan_drug_mas)
pacf(combined_rd_ma_df[combined_rd_ma_df$area == "EU",]$freq_new_orphan_drug_mas)

pacf(combined_rd_ma_df[combined_rd_ma_df$area == "US",]$freq_orphan_drug_mas)
pacf(combined_rd_ma_df[combined_rd_ma_df$area == "EU",]$freq_orphan_drug_mas)

pacf(combined_rd_ma_df[combined_rd_ma_df$area == "US",]$annual_domestic_RD_spending_bil_dollars)
pacf(combined_rd_ma_df[combined_rd_ma_df$area == "EU",]$annual_domestic_RD_spending_bil_dollars)
```


```{r}
sum(combined_rd_ma_df[combined_rd_ma_df$area == "EU",]$freq_orphan_drug_mas)
```


